Exploring the linear separablity of syntactic and semantic information in BERT embeddings

### 2022-06-09 paper/paper.tex
finished dicussion section

We first probe our tasks on the original embeddings, and linear classifers can successfully predict both CCG tags and semantics tags (around 85\% testing accuracy) as resulted in table \ref{role description}. Then we apply INLP method to the linear classifers to derive guarding matrices $P_{syn}$ and $P_{sem}$ which will project the embeddings to the nullspace of syntactic space and semantic space respectively. With some linear tranformation with the original embeddings and the guarding matrices, we are able to extract the embeddings described in table \ref{description}. 
\begin{table}[ht]
    \centering
    \begin{tabular}{p{4cm}p{3cm}}
        expression & description\\ \hline 
        $v_{orig}$ & Original BERT embeddings  \\
        $v_{nosem} = P_{sem} v_{orig}$ & Gained after INLP with semantics task \\
        $v_{nosyn}= P_{syn} v_{orig}$ & Gained after INLP with syntactic task \\
        $v_{sem} = v_{orig}-v_{nosem}$ & Semantics representation \\
        $v_{syn} = v_{orig}-v_{nosyn}$ & Syntactic representation \\
        $v_{syn-sem} = P_{sem} v_{syn}$ &Contains syntactic features only \\
        $v_{sem-syn} = P_{syn} v_{sem}$ &Contains semantic features only \\
        $v_{syn*sem} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{sem}\|}$ & Syntactic representations projected on semantic space\\
        $v_{sem*syn} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{syn}\|}$ & Semantics representations projected on semantic space\\

@@

    Rand($|v_{nosem}|$) & 77 & $84.06\%$ & $87.71\%$ \\
    Rand($|v_{nosyn}|$) & 159 & $82.26\%$ & $87.57\%$ \\
    Rand($|v_{syn-sem}|$) & 81 & p$83.11\%$ & $87.61\%$ \\
    Rand($|v_{sem-syn}|$) & 161 & $82.46\%$ & $86.04\%$ \\
    Rand($|v_{syn*sem}|$) & 0 & p$\%$ & p$\%$ \\
    Rand($|v_{sem*syn}|$) & 0 & p$\%$ & p$\%$ \\
    \hline
    $v_{nosem}$ & 77 & $27.93\%$ & $17.28\%$ \\
    $v_{nosyn}$ & 159 & $23.76\%$ & $49.43\%$ \\
    $v_{syn-sem}$ & 81 & $34.24\%$ & $33.91\%$ \\
    $v_{sem-syn}$ & 161 & $26.96\%$ & $50.21\%$ \\
    $v_{syn*sem}$ & 0 & p$\%$ & p$\%$ \\
    $v_{sem*syn}$ & 0 & p$\%$ & p$\%$ \\
    \hline
    \hline
    \end{tabular}
    \caption{\label{role description} Experiment Result of Different Embeddings
    }
    \end{table*}

@@

\section{Discussion}
\label{sec:discussion}

The INLP method successfully guard information retrieved from probe task. When performing CCG tagging, the model accuracy drops significantly (from 84.06 \% to 23.76\%), and similarly for semantics tagging. To measure the relevancy of semantics information to syntactic task, we compared the performance of CCG tagging with embeddings $v_{nosem}$ with that of embeddings $v_{nosyn}$. The both performances have considerable decrease as to the original embeddings. The model is using representation from both space when probe the syntactic task. Meanwhile, the performances of semantics tagging with the two embeddings differ significantly, with $v_{nosyn}$ performs almost three times better than $v_{nosem}$. The result suggests that the semantics task is less dependent on the syntactic information than the opposite direction, which contradicts the assumption that semantics space will be rel

To investigate the compositionality of semantics space and syntactic space, we also calculated the average cosine simliarity scores between embeddings in table \ref{sim}.

@@

\begin{table}[h]
    \centering
    \begin{tabular}{p{4cm}p{3cm}}
        expression & similarity score\\ \hline 
        $v_{orig}, v_{nosem}$ &  $0.4183$ \\ 
        $v_{orig}, v_{nosyn}$ & $0.3609$\\ 
        $v_{nosyn}, v_{nosem}$ & $0.2830$ \\
        $v_{syn-sem}, v_{sem-syn}$ & $0.2579$ \\

@@

>>>>

We first probe our tasks on the original embeddings, and linear classifers can successfully predict both CCG tags and semantics tags (around 85\% testing accuracy) as resulted in table \ref{role description}. Then we apply INLP method to the linear classifers to derive guarding matrices $P_{syn}$ and $P_{sem}$ which will project the embeddings to the nullspace of syntactic space and semantic space respectively. With some linear tranformation with the original embeddings and the guarding matrices, we are able to extract the embeddings described in table \ref{description}. 
\begin{table}[ht]
    \centering
    \begin{tabular}{p{4cm}p{3cm}}\hline
        \textbf{Expression} & \textbf{Description}\\ \hline 
        $v_{orig}$ & Original BERT embeddings  \\
        $v_{nosem} = P_{sem} v_{orig}$ & Gained after INLP with semantics task \\
        $v_{nosyn}= P_{syn} v_{orig}$ & Gained after INLP with syntactic task \\
        $v_{sem} = v_{orig}-v_{nosem}$ & Semantics representation \\
        $v_{syn} = v_{orig}-v_{nosyn}$ & Syntactic representation \\
        $v_{syn-sem} = P_{sem} v_{syn}$ &Contains syntactic features only \\
        $v_{sem-syn} = P_{syn} v_{sem}$ &Contains semantic features only \\
        $v_{syn*sem} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{sem}\|}$ & Syntactic representations projected on semantic space\\
        $v_{sem*syn} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{syn}\|}$ & Semantics representations projected on semantic space\\

@@

    Rand($|v_{nosem}|$) & 77 & $84.06\%$ & $87.71\%$ \\
    Rand($|v_{nosyn}|$) & 159 & $82.26\%$ & $87.57\%$ \\
    Rand($|v_{syn-sem}|$) & 81 & $83.11\%$ & $87.61\%$ \\
    Rand($|v_{sem-syn}|$) & 161 & $82.46\%$ & $86.04\%$ \\
    Rand($|v_{syn*sem}|$) & 0 & p$\%$ & p$\%$ \\
    Rand($|v_{sem*syn}|$) & 0 & p$\%$ & p$\%$ \\
    \hline
    $v_{nosem}$ & 77 & $27.93\%$ & $17.28\%$ \\
    $v_{nosyn}$ & 159 & $23.76\%$ & $49.43\%$ \\
    $v_{syn-sem}$ & 81 & $34.24\%$ & $33.91\%$ \\
    $v_{sem-syn}$ & 161 & $26.96\%$ & $50.21\%$ \\
    $v_{syn*sem}$ & 0 & p$\%$ & p$\%$ \\
    $v_{sem*syn}$ & 0 & p$\%$ & p$\%$ \\
    \hline
    \hline
    \end{tabular}
    \caption{\label{role description} Experiment Result of Different Embeddings
    }
    \end{table*}

@@

\section{Discussion}
\label{sec:discussion}
The INLP method successfully guard information retrieved from probe task. When performing CCG tagging, the model accuracy drops significantly (from 84.06 \% to 23.76\%), and similarly for semantics tagging. In contrast, when the directions are randomly removed, the performance remains relatively the same for both tasks.

To measure the relevancy of semantics information to syntactic task, we compared the performance of CCG tagging with embeddings $v_{nosem}$ with that of embeddings $v_{nosyn}$. The both performances have considerable decrease as to the original embeddings. The model is using representation from both space when probe the syntactic task. Meanwhile, the performances of semantics tagging with the two embeddings differ significantly, with $v_{nosyn}$ performs almost three times better than $v_{nosem}$. The result suggests that the semantics task is less dependent on the syntactic information than the opposite direction, which contradicts the assumption that syntactic information will be more pertinent to semantics task than vice versa.

We also investigate the compositionality of semantics space and syntactic space, by averaging cosine simliarity scores between embeddings in table \ref{sim}. The simliarity score between $v_orig$ and $v_{nosem}$ is higher than between $v_{orig}$ and $v_{nosyn}$, which might suggest that BERT embeddings contain more syntactic information than semantics. More experiments need to be conducted to verify this assumption.

\begin{table}[h]
    \centering
    \begin{tabular}{p{4cm}p{3cm}}
        \hline
        \textbf{Expressions} & \textbf{Similarity Score}\\ \hline 
        $v_{orig}, v_{nosem}$ &  $0.4183$ \\ 
        $v_{orig}, v_{nosyn}$ & $0.3609$\\ 
        $v_{nosyn}, v_{nosem}$ & $0.2830$ \\
        $v_{syn-sem}, v_{sem-syn}$ & $0.2579$ \\

@@

@@

Our probing experiment 
Though INLP successfully produces the desired guard function, it is worth also worth noting that our dataset is relatively small compared
